
% \section{A Spectrum of GF Grammars for types}
\section{Prior GF Formalizations}

Prior to the grammars explored thesis, Ranta produced two main results
\cite{rantaLog} \cite{aarneHott}. These are incredibly important precedents in
this approach to proof translation, and serve as important comparative work for
which this work responds.

\subsection{CADE 2011}

In \cite{rantaLog}, Ranta designed a grammar which
allowed for predicate logic with a domain specific lexicon supporting mathematical
theories , say geometry or arithmetic, on top of the logic. The syntax was
both meant to be relatively complete, so that typical logical utterances of
interest could be accommodated, as well as support relatively non-trivial linguistic
nuance like lists of terms, predicates, and propositions, in-situ and
bounded quantification, like other ways of constructing more syntactically
nuanced predicates. The more interesting syntactic details captured in this work
was by means of an extended grammar on top of the core. The bidirectional
transformation between the core and extended grammars via a PGF also show the
viability and necessity of using more expressive programming languages (Haskell) when
doing thorough translations.
 
Lists are natural to humans - this is reflected in our language. The RGL supports listing the sentences, noun phrases, and other
grammatical categories. One can then use PGF to unroll the lists into binary
operators, or alternatively transform them in the opposite direction.
, we first mention that GF
natively supports list categories, the judgment \term{cat [C] {n}} can be
desugared to
\begin{verbatim}
  cat ListC ;
  fun BaseC : C -> ... -> C -> ListC ; -- n C ’s
  fun ConsC : C -> ListC -> ListC
\end{verbatim}

As a case study for this grammar, the proposition $\forall x (Nat(x) \supset
Even(x) \lor Odd(x))$ can be given a maximized and minimized version. The tree
representing the \emph{syntactically complete} phrase ``for all natural numbers
x, x is even or x is odd" would be minimized to a tree which linearizes to the
\emph{semantically adequate} phrase ``every natural number is even or odd".

% The trees below also revels how the semantically adequate tree is also
% simpler to understand as representing
% \begin{verbatim}
% PUnivs
%   (BaseVar X)
%   Nat
%   (PConj COr (PAtom (APred1 Even (IVar X))) (PAtom (APred1 Odd (IVar X))))
% |
% V
% PAtom (APred1 (ConjPred1 COr (BasePred1 Even Odd)) (IUniv Nat))
% \
% end{verbatim}

We see that our criteria of semantic adequacy and syntactic completeness can
both occur in the same grammar, with the different subsets related not by a
direct GF translation but a PGF level transformation. Problematically, this
syntactically complete phrase produces four ASTs, with the ``or" and ``forall"
competing for precedence. Where PGF may only give one translation to the
extended syntax, this doesn't give the user of the grammar confidence that her
phrase was correctly interpreted.

In the opposite direction, the desugaring of a logically ``informal"
statement into something less linguistically idiomatic is also accomplished.
Ranta claims ``Finding extended syntax equivalents for core syntax trees is
trickier than the opposite direction". While this may be true for this
particular grammar, we argue that this may not hold generally true, for a few
reasons. Dealing with these ambiguities must be
solved first and foremost to satisfy the PL designer who only accepts
unambiguous parses. For instance, the gf shell shows \codeword{p "the sum of the
sum of x and y and z is equal to the sum of x and the sum of y and z"} giving 32
unique parses. While these can be filtered through some simple modifications, we
should note that a mapping a complicated Agda expression to a bigger extended syntax 
would likely be infeasible. Ranta also outlines the mapping, $\llbracket -
\rrbracket : Core \to Extended$,

\begin{itemize}
\item Flattening a list \\
  $x\ and\ y\ and\ z\ \mapsto x,\ y\ and\ z$
\item Aggregation \\
  $x\ is\ even\ or\ x\ is\ odd\ \mapsto x\ is\ even\ or\ odd$
\item In-situ quantification \\
  $\forall\ n\ \in Nat,\ x\ is\ even\ or\ x\ is\ odd \mapsto every\ Nat\ is\ even\ or \odd$
\item Negation \\
  $it\ is\ not\ that\ case\ that\ x\ is\ even\ \mapsto \x is\ not\ even$
\item Reflexivitazion \\
  $x\ is\ equal\ to\ x\ \mapsto x\ is\ equal\ to\ itself$
\item Modification \\
  $x\ is\ a\ number\ and\ x\ is\ even\ \mapsto x\ is\ an\ even\ number$
\end{itemize}

Scaling this to cover more phenomena, as outlined in [cite ganesalingam], will
pose challenges. Extending this work in general without very sophisticated
statistical methods is impossible because mathematicians will speak uniquely,
and so choosing how to build an extended that covers multiple ways of saying
things will be require many arbitrary choices along the way, and efficient
communication is a pragmatic matter outside the scope of this work. The most
interesting linguistic phenomena, we believe, is in-situ quantification, in part
because quantifiers been at the epicenter of the Montague tradition.

In some sense, this grammar serves as a case study for what this thesis is
trying to do. However, we note that the core logic only supports propositions
without proofs - it is not a type theory with terms. Additionally, the domain of
arithmetic is an important case study, but scaling this grammar (or any other,
for that matter) to allow for \emph{semantic adequacy} of real mathematics is
still far away, or as Ranta concedes, ``it seems that text generation involves
undecidable optimization problems that have no ultimate automatic solution." It
would be interesting to further extend this grammar with both terms and an
Agda-like concrete syntax.

\subsubsection{An Additional Grammar}

One of the difficulties encountered in this thesis was back-tracing previously
written code - the large size of a grammar and declarative nature of the code,
makes it incredibly difficult to isolate individual features one may wish to
understand. This is true for both GF and PGF, and therefore a lot of work went
into filtering grammars so that one could just try to isolate a single feature
of interest. Although the GF module system is partially capable of solving this
in some circumstances, there is not proper methodology and therefore it was
found that writing a grammar from scratch was often the easiest way to do this.
Grammars can be written compositionally (adding new categories and functions, refactoring
linearization types, etc.) but decomposing them is not a compositional process. 

Additionally, testing the intermediary functions in the PGF grammar was, at
least initially, quite difficult, although once one becomes familiar with the
PGF API there is a possibility of streamlining at least parts of this process.

We wrote a smaller version [cite mycode] of, just focused on propositional
logic, but with the added interest of not just translating between Trees, but
also allowing Haskell computation and evaluation of expressions. Although this
exercise was in some ways a digression from the language of proofs, it also
highlighted many interesting problems.  

We begin with an example : the idea was to create a PGF layer for the evaluation
of propositional expressions to their Boolean values, and then create a question
answering system which gave different types of answers - the binary valued
answer, the most verbose possible answer, and the answer which was deemed the
most semantically adequate, \codeword{Simple}, \codeword{Verbose}, and
\codeword{Compressed}, respectively. The system is capable of the following :
\begin{verbatim}
is it the case that if the sum of 3 , 4 and 5 is prime , odd and even then 4 is prime and even
  Simple : yes .
  Verbose : yes . if the sum of 3 and the sum of 4 and 5 is prime and the sum of 3 and the sum of 4 and 5 is odd and the sum of 3 and the sum of 4 and 5 is even then 4 is prime and 4 is even .
  Compressed : yes . if the sum of 3 , 4 and 5 is prime , odd and even then 4 is prime and even .
\end{verbatim}

The extended grammar in this case only had lists of propositions and predicates,
and so it was much simpler than in [cite logic]. GF list categories are then
transformed into Haskell lists, so the syntactic sugar for a GF list is actually
functionally tied to its external behavior as well. The functions for our
discussion are:
\begin{verbatim}
  IsNumProp : NumPred -> Object -> Prop ;
  LstNumPred : Conj -> [NumPred] -> NumPred ; 
  LstProp : Conj  -> [Prop] -> Prop ;
\end{verbatim}

Note that a numerical predicate, \codeword{NumPred}, represents, for instance, primality.
In order for our pipeline to answer the
question, we had to not only do transform trees, $\llbracket - \rrbracket :
\{pgfAST\} \rightarrow \{pgfAST\}$
, but also evaluate them in more classical domains $\llbracket - \rrbracket : \{pgfAST\} \rightarrow
\mathds{N}$ for the arithemtic objects and $\llbracket - \rrbracket : \{pgfAST\} \rightarrow
\mathds{B}$ for the propositions, the later case being  \codeword{evalProp} below.

Due to the extension, there are now more tricky cases one has to cover when
evaluating this, which for a normal ``propositional evaluator" doesn't have to
deal with lists. For the most part, this evaluation is able to just apply
boolean semantics to the \emph{canonical} constructors, like GNot. However, a
bug that was subtle and difficult to find appeared, thereby forcing us to dig
deep inside GIsNumProp, preventing an easy solution to what would otherwise be a
simple example of denotational semantics.
\begin{verbatim}
evalProp :: GProp -> Bool
evalProp p = case p of
  ...
  GNot p -> not (evalProp p)
  ...
  GIsNumProp (GLstNumProp c (GListNumPred (x : xs))) obj ->
    let xo = evalProp (GIsNumProp x obj)
        xso = evalProp (GIsNumProp (GLstNumProp c (GListNumPred (xs))) obj) in
    case c of
      GAnd -> (&&) xo xso 
      GOr -> (||) xo xso
  ... 
\end{verbatim}
While still relatively tame to overcome, one can imagine that an even expressive and semantically
adequate abstract syntax might yield many trickier and more subtle obstacles like this.
The more semantic content one incorporates into the GF grammar, the larger the
PGF GADT, which leads to many more pains when evaluating these trees.


Personally speaking, there were many obstacles in getting a relatively simple
example to work, particularly when it came to writing test cases. For the naive
way to test with GF is to translate, and the linearization and parsing functions
don't give the programmer many degrees of freedom. While some work has been
done [cite inari] in allowing for testing of grammars. The specific domain of
formal languages in GF require a more refined notion of testing because
they should be testable relative to some model with well behaved mathematical
properties.



Debugging something in pipeline $String \rightarrow GADT \rightarrow GADT
\rightarrow String$ for a large scale grammar without a testing methodology for
each intermediate state is surely to be avoided. Unfortunately, there is no
published work on using Quickcheck [cite hughes] with PGF. The bugs in this
grammar were discovered via the input and output \emph{appearance} of strings .
Often, no string would be returned, and discovering the source was excruciating.
In one case, a bug was discovered that was presumed to be from the PGF
evaluator, but was then back-traced to Ranta's grammar from which the code had
been refactored. The sentence which broke our pipeline from core to extended, "4
is prime , 5 is even and if 6 is odd then 7 is even", would be easily
generated through something like quickcheck.  

An important observation that was made during this development : that theorems
should be the source of inspirations for deciding which PGF transformations
should take place. For instance, one could say, define $odd : \mathds{N}
\rightarrow Set$, $prime : \mathds{N} \rightarrow Set$ and prove that $\forall n
\in \mathds{N}.\; n > 2 \times prime\; n \implies odd\; n$. We can use this
theorem as a source of translation, and in fact encode a PGF rule that
transforms anything of the form ``n is prime and n is odd" to ``n is prime",
subject to the condition that $n \neq 2$. One could then take a whole set of
theorems from predicate calculus and encode them as Haskell functions which
simplify the expressions to give some kind of adequate kernel of the expression.
The verbose ``if $a$ then $b$ and if $a$ then $c$, should always be read ``if
$a$ then $b$ and $c$". see The application of these theorems and evaluation
functions in Haskell could help give in our QA example, more informative and
direct answers.

We hope this intricate look at a fairly simple grammar highlights some very
serious considerations one should make when writing a PGF embedded grammar.
These include : how does the semantic space the grammar seeks to approximate
effects the PGF translation, how testing formal grammars is non-trivial but
necessary future work, and finally, how information (in this case theorems) from
the domain of approximation can shape and inspire the PGF transformations for
during the translation process.


% change this to section, add something about PGF above
% also show agda code which reflects difficulty with universes
% discuss mixture of logic language, tt, and sets in the example chosen, and the
% fact that despite its successes, 
% contrast with hott book which is exclusively meant to be type-theoreticly formulatedwritten
% formal, personal, and sociological problems, for conclusion
% compare and contrast the categories alone, how they are similair and different
% after all three have been written
%IN ORDER TO USE GF TO MODEL FORMAL LANGUAGES, ONE HAS TO DEVELOP FORMAL MODELS
%OF GF ITSELF. this includes testing, extracting grammars, proving that GF
%grammar does or doesn't contain a string, how to control ambiguity with
%parsing. also, look at notions of subgrammars (and sublanguages), with the
%module systems, and with respect to
% can we come up with a model of computation for GF (i.e. lambdas, Strings, etc)?

\subsection{Stockholm Math Seminar 2014}

In 2014, Ranta gave an unpublished talk at the Stockholm Mathematics seminar
\cite{aarneHott}. Fortunately the code is available, although many of the design
choices aren't documented in the grammar. This project aimed to provide a
translation like the one desired in our current work, but it took a real piece
of mathematics text as the main influence on the design of the Abstract syntax.

This work took a page of text from Peter Aczel's book which more or less goes
over standard HoTT definitions and theorems, and allows the translation of the
latex to a pidgin logical language. The central motivation of this grammar was
to capture, entirely ``real" natural language mathematics, i.e. that which was
written for the mathematician. Therefore, it isn't reminiscent of the slender
abstract syntax the type theorist adores, and sacrificed ``syntactic
completeness" for ``semantic adequacy". This means that the abstract syntax is
much larger and very expressive, but it no longer becomes easy to reason about
and additionally quite ad-hoc. Another defect is that this grammar
overgenerates, so producing a unique parse from the PL side will become tricky.
Nonetheless, this means that it's presumably possible to carve a subset of the
GF HoTT abstract file to accommodate an Agda program, but one encounters rocks as soon
as one begins to dig. For example, in \autoref{fig:M1} is some rendered latex
taken verbatim from Ranta's test code.

With some of hours of tinkering on the pidgin logic concrete syntax and some
reverse engineering with help from the GF shell, one is able to get these
definitions in \autoref{fig:M2}, which are intended to share the same syntactic
space as cubicalTT. We note the first definition of ``contractability" actually
runs in cubicalTT up to renaming a lexical items, and it is clear that the
translation from that to Agda should be a benign task. However, the
\emph{equivalence} syntax is stuck with the artifact from the bloated abstract
syntax for the of the anaphoric use of ``it", which may presumably be fixed with
a few hours more of tinkering, but becomes even more complicated when not just
defining new types, but actually writing real mathematical proofs, or relatively
large terms. To extend this grammar to accommodate a chapter worth of material,
let alone a book, will not just require extending the lexicon, but encountering
other syntactic phenomena that will further be difficult to compress when
defining Agda's concrete syntax.

\input{latex/ContrClean}

% \begin{figure}

%  \textbf{Definition}:
%  A type $A$ is contractible, if there is $a : A$, called the center of contraction, such that for all $x : A$, $\equalH {a}{x}$.

%  \textbf{Definition}:
%  A map $f : \arrowH {A}{B}$ is an equivalence, if for all $y : B$, its fiber, $\comprehensionH {x}{A}{\equalH {\appH {f}{x}}{y}}$, is contractible.
%  We write $\equivalenceH {A}{B}$, if there is an equivalence $\arrowH {A}{B}$.
% \caption{Rendered Latex} \label{fig:M1}


% \begin{verbatim}
% isContr ( A : Set ) : Set = ( a : A ) ( * ) ( ( x : A ) -> Id ( a ) ( x ) )

% Equivalence ( f : A -> B ) : Set = 
%   ( y : B ) -> ( isContr ( fiber it ) ) ; ; ; 
%   fiber it : Set = ( x : A ) ( * ) ( Id ( f ( x ) ) ( y ) )
% \end{verbatim}
% \caption{Pidgin cubicalTT} \label{fig:M2}

% \input{latex/ContrClean}
% % \input{latex/ex}
% % \input{latex/primitives}

% \caption{Agda} \label{fig:M3}
% \end{figure}

Additionally, we give the Agda code in \autoref{fig:M3}, so-as to see what the
end result of such a program would be. The astute reader will also notice a
semantic in the pidgin rendering error relative to the Agda implementation.
\codeword{fiber} has the type \codeword{it : Set} instead of something like
\codeword{(y : B) : Set}, and the y variable is unbound in the \codeword{fiber}
expression. This demonstrates that to design a grammar prioritizing
\emph{semantic adequacy} and subsequently trying to incorporate \emph{syntactic
completeness} becomes a very difficult problem. Depending on the application of
the grammar, the emphasis on this axis is most assuredly a choice one should
consider up front.

While both these grammars have their strengths and weaknesses, one shall see
shortly that the approach in this thesis, taking an actual programming language
parser in Backus-Naur Form Converter (BNFC), GFifying it, and trying to use the
abstract syntax to model natural language, gives in some sense a dual challenge,
where the abstract syntax remains simple, but its linearizations become
must increase in complexity.


% below is prior text, probably discard

% We now discuss the various iterations of code which experimented with NL aspects

% We should again emphasize the role of, in particular, Rantas two grammars, one
% formalizing logic, and the other working with a case study of a real text\cite{aarneHott}



% We now discuss the GF implementation, capable of parsing both natural language
% and Agda syntax. The parser was appropriated from the cubicaltt BNFC parser,
% de-cubified and then gf-ified. The languages are tightly coupled, so the
% translation is actually quite simple. Some main differences are:

% \begin{itemize}[noitemsep]

% \item GF treats abstract and concrete syntax seperately. This allows GF to
% support many concrete syntax implementation of a given grammar

% \item Fixity is dealt with at the concrete syntax layer in GF.  This allows for
% more refined control of fixity, but also results in difficulties : during
% linearization their can be the insertion of extra parens.

% \item GF supports dependent hypes and higher order abstract syntax, which makes
% it suitable to typecheck at the parsing stage. It would very interesting to see
% if this is interoperable with the current version of this work in later
% iterations [Todo - add github link referncing work I've done in this direction]

% \item GF also is enhanced by a PGF back-end, allowing an embedding of grammars
% into, among other languages, Haskell.

% \end{itemize}

% While GF is targeted towards natural language translation, there's nothing
% stopping it from being used as a PL tool as well, like, for instance, the
% front-end of a compiler. The innovation of this thesis is to combine both uses,
% thereby allowing translation between Controlled Natural Languages and
% programming languages.

% Example expressions the grammar can parse are seen below, which have been
% verified by hand to be isomorphic to the corresponding cubicaltt BNFC trees:

% \begin{verbatim}

% data bool : Set where true | false 
% data nat : Set where zero | suc ( n : nat )  
% caseBool ( x : Set ) ( y z : x ) : bool -> Set = split false -> y || true -> z
% indBool ( x : bool -> Set ) ( y : x false ) ( z : x true ) : ( b : bool ) -> x b = split false -> y || true  -> z
% funExt  ( a : Set )   ( b : a -> Set )   ( f g :  ( x : a )  -> b x )   ( p :  ( x : a )  -> ( b x )   ( f x ) == ( g x )  )  : (  ( y : a )  -> b y )  f == g = undefined
% foo ( b : bool ) : bool = b

% \end{verbatim}

% [Todo] add use cases

