\subsection{Agda}

\subsubsection{Overview}
Agda is an attempt to faithfully formalize Martin-Löf's intensional type theory
\cite{ml1984}. Referencing our previous distinction, one can think of
Martin-Löf's original work as a specification, and Agda as one possible
implementation.

Agda is a functionaly programming language which, through an interactive
environment, allows one to iteratively apply rules and develop constructive
mathematics. It's current incarnation, Agda2 (but just called Agda), was
preceded by ALF, Cayenne, and Alfa, and the Agda1. On top of the basic MLTT,
Agda incorporates dependent records, inductive definitions, pattern matching, a
versatile module system, and a myriad of other bells and whistles which are of
interest generally and in various states of development but not relevant to this
work.

For our purposes, we will only look at what can in some sense be seen as the
kernel of Agda. Developing a full-blown GF grammar to incorporate more
advanced Agda features would require efforts beyond the scope of this work.

Agda's purpose is to manifest the propositions-as-types paradigm in a practical
and useable programming language. And while there are still many reasons one may
wish to use other programming languages, or just pen and paper to do her work,
there is a sense of purity one gets when writing Agda code. There are many good
resources for learning Agda \cite{Bove2009} \cite{stump} \cite{ulf}
\cite{wadler} so we'll only give a cursory overview of what is relevant for this
thesis, with a particular emphasis on the syntax.

\input{latex/ex}

\input{latex/twin-primes}

